Species richness destabilizes trophic communities in freshwater ecosystems

### 2020-04-01 manuscript/bef_stability/manuscript.Rmd
Update Introduction + Biomass, protocol and variability methods

---
title: Species richness destabilizes trophic communities in freshwater ecosystems
author: Alain Danet, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today 
output:
  pdf_document:
    fig_caption: true 
    keep_tex: true
fontsize: 12pt
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
geometry: margin=2cm
bibliography: references.bib
---

@@

##### BEF
Determining the shape of the relationship between species diversity and ecosystem
functioning has been a central research topic in ecology
[@may_will_1972, @hector_plant_1999, @loreau_biodiversity_2001, @hooper_effects_2005, @duffy_why_2009, @isbell_forest_quantifying_2018].
Current biodiversity losses
in all ecosystems challenge scientists to understand how ecosystems will be
affected in turn. Studies in Bioversity-Ecosystem Functioning (BEF) usually
quantify species diversity as species richness (i.e number of species) while
ecosystem functioning is quantified as total biomass, productivity (biomass
produced by unit of time) or stability which is quantified in more diverse ways
such as the variability of species abundance, the resilience to perturbation, or
resistance to perturbation [@donohue_navigating_2016].

Species richness has been reported to increase productivity in grasslands
[@grace_integrative_2016], but complex links between between species richness,
environment, productivity and total biomass showed that species richness
decrease total biomass [@grace_integrative_2016] contrary to previous
assessment [@hector_plant_1999, @tilman_biodiversity_2014]. Complementarity
between species in resource exploitation [@cardinale_species_2002] leading to
overyielding (CHECK) has been proposed to explain positive relationship between
productivity and species richness, while species richness effect can also rise
from the statistical probability to end up with a highly productive species
. The relationship between species richness and stability is even more complex
partly because of the inherent multidimensionality of stability
[@donohue_dimensionality_2013, @donohue_navigating_2016] and because of the
different metrics of stability privileged in theoretical (stability of local
equilibrium, robustness to primary extinction) and empirical work (variability
of abundance or biomass through time). @may_will_1972 stimulated a lot of
debate and fecund research, showing that species richness, probability of
interacting and average interaction strength destabilizes communities in which
species interact randomly. Empirically, consistent works showed that species
richness increased community stability in grasslands
[@tilman_biodiversity_2006], mesocosms [@pennekamp_biodiversity_2018] and
vertebrates [@olivier_independent_nodate, @franssen_annual_2011]. Two
components can explain variation of community stability, (1) individual species
stability and (2) the species synchrony [@loreau_species_2008, @thibaut_understanding_2013].
Positive [@tilman_biodiversity_2006], neutral
[@jiang_different_2009, @olivier_independent_nodate] and negative
[@franssen_annual_2011, @olivier_independent_nodate] relationships between
species stability and species richness have been found. On the other side,
studies consistently found that species richness decreased synchrony between
species [@loreau_species_2008, @thibaut_understanding_2013, @bluthgen_land_2016; @olivier_independent_nodate]. However, as far as we know,
less studies have reported empirical assessment of stability-richness
relationship in trophic network [but see @thebault_stability_2010].

Conversely, an extensive literature exists on the effects of network structure
on stability in mutualistic and trophic networks
[@angelis_stability_1975, @stouffer_, @thebault_stability_2010, @barbier_pyramids_2019, @shanafelt_david_w._stability_nodate, add ref].
Particularly, in trophic
networks, most studies are theoretical and consider asymptotic stability to
small perturbation. Studies showed that compartimentalization [@stouffer_] can
increase persistence, while nestedness can either increase
[@bascompte, @thebault_stability_2010] or decrease persistence
[@thebault_stability_2010].
Connectance can decrease [@thebault_stability_2010] or increase
[@angelis_stability_1975] resilience of trophic network depending on the
functioning of networks [@angelis_stability_1975].

---
#- Vertical diversity and stability/biomass
  #- well studied: theoretically and (experimentally, less)
  #- mecanisms:
    #- stability: top-down and bottom-up effect
    #- biomass: top-down and bottom-up effects (pyramids of biomass)
  #- key papers:[@barbier_pyramids_2019, @shanafelt_david_w._stability_nodate]
  #- BUT: local stability, little systems 
---

Trophic network are assembled both vertically and horizontally. Studies showed
that the vertical position in the trophic network as well as the strength of
top-down and bottom-up regulation determine biomass and stability at
equilibrium  [@shanafelt_david_w._stability_nodate, @barbier_pyramids_2019].
However studies on stability-diversity have focused on either vertical
diversity or horizontal but rarely both. @zhao_horizontal_2019 have shown that
horizontal and vertical diversity respectively increase and decrease stability
in plankton communities.

---
#- Big scope assessment:
  #- inconsistency between theoretical and empirical stability
    #- theoretical: no cv_sp nor synchrony
  #- No studies on stability-richness in empirical trophic communities
    #- no effect of trophic structure
---

A large inconsistency remains between theoretical and empirical assessment of
stability-diversity in trophic communities. Theoretical studies have
ignored temporal variability and the relative contribution of synchrony and
species variability while empirical studies about temporal variability in natural 
communities have ignored trophic communities and the effect of trophic structure.

---
#- Goal: Assess relation biomass/stability-diversity in empirical trophic communities
  #- introduce effect of trophic structure and environment
  #- merge gap between theoretical and empirical studies
---

---
#- What we have done:
  #- big empirical dataset
  #- rebuilding empirical network
  #- assess relationship between environment, community structure and stability
---

Our goal was to assess stability-diversity in natural trophic community with the
effect of the trophic structure. We combined together the effect of vertical
and horizontal diversity and thus aimed to bridge the gap between theoretical
and empirical studies. We used an standardized monitoring of freshwater fish
communities in french streams. More than three hundreds sites were followed
more than 10 years on the 1995-2018 period. We inferred four thousands potential
interaction networks corresponding to each sampling event, based on size
relationships and knowledge about their ontogenic diet shift.

@@

We investigated the effects of the community structure on total biomass and on
synchrony and species variability, which in turn determine community
variability. We have also included the effect of environmental drivers which
can directly affect community variability or indirectly via community structure.
Finally, we assessed the effect of horizontal diversity on trophic group
biomass and variability.

---
#- Environment can drive stability and biomass:
  #- forcing: environmental forcing on communities 
  #- can lead to misleading effect: @grace_integrative_2016
---

@@

## Fish data

@@

Fish communities were monitored across streams of France on the period
1995-2018 by a French governmental agency (OFB, previously AFB and ONEMA). We
kept only standardized protocols: complete and partial methods for respectively
little and big streams (detailed in Supplementary methods).  The sampling was
performed by electric fishing. All the stunned fish species were identified and
the individuals counted. Fish were classified in batch according to species and
size class. Depending on the size class, individual fishes were measured or
only a subsample. The size of individual fishes in subsampled batches were
inferred under normality distribution assumption (detailed in Supplementary
material).

@@

```{r}
myload(metaweb_analysis, dir= mypath("data"))
meta <- metaweb_analysis
#metaweb_analysis$resource
#meta$metaweb[meta$resource, meta$resource]

@@

The body mass of each fish was approximated from their body size with an
allometric law ($B = 0.01 \times (l^3.03)$, $B$ being the body mass in grams and $l$ the
body size in millimeters). We have worked with body mass instead of abundances to avoid to sum
abundances of fishes that had really different sizes and so body masses. Body
mass are more meaningful in terms of ecosystem description. The biomass of each
trophic species (and species) was divided by the surface sampled (measured in
square meters) during the sampling event. The total biomass in a given
community are so expressed in $gm^{-2}$.

@@

The biomass stability was decomposed in synchrony and weighted average CV
components according to @thibaut_understanding_2013:

>>>>

---
title:"""
  Complex mecanisms linking biomass stability and total biomass with
species richness and network structure in stream ecosystems 
"""
author: Alain Danet, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today 
output:
  pdf_document:
    fig_caption: true 
    keep_tex: true
fontsize: 12pt
header-includes:
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
geometry: margin=2cm
bibliography: references.bib
---

@@

#### Importance of BEF assessment

Determining the shape of the relationship between species diversity and ecosystem
functioning has been a central research topic in ecology
[@may_will_1972, @hector_plant_1999, @loreau_biodiversity_2001, @hooper_effects_2005, @duffy_why_2009, @isbell_forest_quantifying_2018].
Current biodiversity losses
in all ecosystems challenge scientists to understand how ecosystems will be
affected in turn. Studies in Bioversity-Ecosystem Functioning (BEF) usually
quantify species diversity as species richness (i.e number of species) while
ecosystem functioning is quantified as total biomass, productivity (biomass
produced by unit of time) or stability which is quantified in more diverse ways
such as the variability of species abundance, the resilience to perturbation, or
resistance to perturbation [@donohue_navigating_2016].

#### Shape of BEF: total biomass

Species richness has been reported to increase productivity
[@tilman_productivity_1996, @grace_integrative_2016] through complementarity
between species in resource exploitation and selection effects
[@loreau_partioning_2001, @cardinale_species_2002, @fargione_from_2007].
Empirical assessment of the effect of species richness on total biomass showed negative
[@grace_integrative_2016] and positive relationships [@hector_plant_1999,
@tilman_biodiversity_2014]. @grace_integrative_2016 showed that the results
depend on the refinement of the mecanisms involved in the analysis.

#### Shape of BEF: stability

The relationship between species richness and stability is even more complex
partly because of the inherent multidimensionality of stability
[@donohue_dimensionality_2013, @donohue_navigating_2016] and because of the
different metrics of stability privileged in theoretical (stability of local
equilibrium, robustness to primary extinction) and empirical work (variability
of abundance or biomass through time). Empirical studies showed consistently
that species richness decreased community variability in grasslands
[@tilman_biodiversity_2006], mesocosms [@pennekamp_biodiversity_2018] and
vertebrates [@olivier_independent_nodate, @franssen_annual_2011]. Two
mecanisms are involved into community variability, (1) species
variability and (2) the species synchrony [@loreau_species_2008, @thibaut_understanding_2013].
Positive [@tilman_biodiversity_2006], neutral
[@jiang_different_2009, @olivier_independent_nodate] and negative
[@franssen_annual_2011, @olivier_independent_nodate] relationships between
species variability and species richness have been found. On the other side,
studies consistently found that species richness decreased synchrony between
species
[@loreau_species_2008, @thibaut_understanding_2013, @bluthgen_land_2016, @olivier_independent_nodate].
Beyond species richness, species interactions, which modifies species
abondance and biomass and make species dynamics interdependant, have been
largely ignored in empirical assessment.

#### BEF: Theoretical ecology

Conversely, theoretical ecology display an extensive literature about the
effects of network structure on stability 
[@may_will_1972, @angelis_stability_1975, @stouffer_compartmentalization_2011, @thebault_stability_2010, @barbier_pyramids_2019, @shanafelt_david_w._stability_nodate].
Theoretical ecology generally uses stability definitions related to community
equilibrium and its stability to small perturbations.
Connectance (i.e. proportion of realized links in a network) can decrease
[@angelis_stability_1975, @thebault_stability_2010] or increase resilience of
trophic network depending on the functioning of networks
[@angelis_stability_1975].
The number of trophic levels, the vertical position in the trophic network as
well as the strength of their biomass regulation (i.e. top-down and bottom-up)
can determine biomass and stability at equilibrium
[@hurd_stability_1971, @shanafelt_david_w._stability_nodate, @zhao_horizontal_2019, @barbier_pyramids_2019].

A large inconsistency remains between theoretical and empirical assessments of
the relationship between community structure and ecosystem properties.
Theoretical studies have ignored temporal variability and then the relative
contribution of synchrony and species variability on community variability while
empirical studies about temporal variability and total biomass in natural
communities have ignored the contribution of network structure.

Our goal was to assess the mecanisms determining temporal variability and total
biomass in natural communities while integrating both species richness and
network structure. We combined together the effect of vertical
and horizontal diversity and thus aimed to bridge the current gap between theoretical
and empirical studies. We used an standardized monitoring of fish
communities in french streams. More than three hundreds sites were followed
more than 10 years on the 1995-2018 period. We inferred four thousands potential
interaction networks corresponding to each sampling event, based on trait
matching.

@@

We investigated the effects of the community structure on total biomass and
community variability, which in turn determine community variability. We have
also included the effect of environmental drivers [@grace_integrative_2016]
which can directly affect community variability or indirectly via community
structure.

---
#- Environment can drive stability and biomass:
  #- forcing: environmental forcing on communities 
  #- can lead to misleading effect: 
---

@@

## Community variability

@@

#### Protocol

Fish communities were monitored across streams of France on the period
1995-2018 by a French governmental agency (OFB, previously AFB and ONEMA). We
kept only standardized protocols: complete and partial methods for respectively
little and big streams (detailed in Supplementary methods). In the complete
protocol, the sampling takes place over the whole width of the river and is done
by foot whereas in the partial protocol, the sampling is done over the bank only
and by boat. The sampling took place either in late spring and/or Autumn.
Heterogeneity in the sampling time may introduce variability that is not related
to ecological processes such as the presence of juveniles. In order to reduce
heterogeneity in sampling time, we select only one sampling per year if there
was two in a given year. Most of the fish sampling occurred in September
(Median: mid-August) and within station the median standard deviation of the
sampling month was inferior to one (0.7). Additionally, we removed sampling
events, in a given station, which the length of the stream sampled was outside
the median length sampled more or less 30%. The number of sampling events is
critical in order to estimate a meaningful variability of the community. We kept
only station that have been followed ten years or more (i.e 10 sampling events
or more), which sums to 403 stations.

The sampling itself was performed by electric fishing. All the stunned fish
species were identified and the individuals counted.

##### Biomass

To characterize communities variability, abundance or biomass are generally
used. When the body size and therefore body mass vary by several orders of
magnitude, one downside of abundance is that we mix individual that have
potentially really different meaning in terms of ecosystem properties such as
biomass. We therefore estimated the body mass of fish individuals according to
their body size. During sampling, the fishes were classified in batch according
to species and size class. Depending on the batches, the body size of all
individuals were measured or only a subsample. Only the body size big fishes was
measured individually. The body size of individual fishes in subsampled batches
was inferred under normality distribution assumption (detailed in Supplementary
material). The body mass of individual fishes were then estimated with an
allometric law ($B_i = 0.01 \times l_i^3.03$, $B$ being the body mass in grams,
$l$ the body size in millimeters and $i$ the individual).

The species and community biomass were then computed for each sampling event.
As sampling effort varied across protocol type and sampling event, biomass
was reported to the surface sampled (in $m^{2}$). Community and species biomass
were then expressed in $gm^{-2}$.

Total biomass was characterized as the median total biomass over time. Community
stability was defined as the inverse of biomass variability. Variability was
computed as Coefficient of Variation (CV) of the biomass ($CV_i = \sigma_i \mu_i^{-1}$,
with $\sigma_i$ and $\mu_i$ being respectively the standard deviation and the
average of the biomass of the station $i$). Then, biomass stability was equal to
$CV^{-1} = \mu \sigma^{-1}$.

The CV of community biomass can represent very different dynamics such as
decreasing, stable or increasing. However, the concept of community variability
refered to variability around an equilibrium, which is rarely observed or
difficult to characterize in natural communities. 
We However used SAX (Symbolic Aggregate Approximation) method to caracterize
rougly biomass dynamics over time (details in Supplementary information). We
finally kept only the stations which the biomass had no trends, remaining 99
stations.

To characterize the mecanisms driving biomass stability, we decomposed
the community CV in synchrony and weighted average species CV
components according to @thibaut_understanding_2013:

Species richness was computed as the total species richness over time in a given
station. To normalize by sampling effort, we reported total species richness to
the sum of surface sampled over the time period, it was expressed in number of
species by square meter.

@@

```{r}
myload(metaweb_analysis, dir= mypath("data"))
meta <- metaweb_analysis
#metaweb_analysis$resource
#meta$metaweb[meta$resource, meta$resource]
```

@@

@@
